XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-24 talk/icooolps2011/paper.tex
use Verbatim environments
reshuffle some stuff, kill one of the traces

@@

%___________________________________________________________________________
\section{The PyPy Project}
\label{sect:pypy}

@@

%___________________________________________________________________________
\section{Tracing JIT Compilers}
\label{sect:tracing}

XXX

%___________________________________________________________________________
\section{Controlling The Extent of Tracing}

\subsection{Background}

First, let's recap some basics: PyPy's approach to implementing dynamic
languages is to write an interpreter for
the language in RPython. This interpreter can be translated to C and then
further to machine code. The interpreter consists of code in the form of a
large number of generated C functions and some data. Similarly, the user
program consists of functions in the language the interpreter executes.

XXX As was explained in a \href{http://morepypy.blogspot.com/2009/03/applying-tracing-jit-to-interpreter.html}{blog post} and a \href{http://codespeak.net/svn/pypy/extradoc/talk/icooolps2009/bolz-tracing-jit.pdf}{paper} two years ago, PyPy's JIT is a
meta-tracer. Since we want to re-use our tracer for a variety of languages, we
don't trace the execution of the user program, but instead trace the execution
of the \emph{interpreter} that is running the program. This means that the traces
don't contain the bytecodes of the language in question, but RPython-level
operations that the interpreter did to execute the program.

@@

@@

The last section described how to control the
extent of tracing. In this section we will describe how to add hints that
influence the optimizer.  If applied correctly these techniques can give
really big speedups by pre-computing parts of what happens at runtime. On the other
hand, if applied incorrectly they might lead to code bloat, thus making the
resulting program actually slower.

\subsection{Background}

Before sending the trace to the backend to produce actual machine code, it is
optimized.  The optimizer applies a number of techniques to remove or reduce
the number of operations: most of these are well known \href{http://en.wikipedia.org/wiki/Compiler_optimization\#Optimization_techniques}{compiler optimization
techniques}, with the difference that it is easier to apply them in a tracing
JIT because it only has to deal with linear traces.  Among the techniques:
%
\begin{itemize}
    \item constant folding
    \item common subexpression elimination
    \item allocation removal \cite{bolz_allocation_2011}
    \item store/load propagation
    \item loop invariant code motion
\end{itemize}

In this section we describe two hints that allow the interpreter author to
increase the optimization opportunities for constant folding. For constant
folding to work, two conditions need
to be met:
%
\begin{itemize}
    \item the arguments of an operation actually need to all be constant,
    i.e. statically known by the optimizer
    \item the operation needs to be \emph{pure}, i.e. always yield the same result given
    the same arguments.
\end{itemize}

@@

\textbf{Note}: These hints are written by an interpreter developer and applied to the
RPython source of the interpreter. Normal Python users will never see them.

@@

If the fragment is traced with \texttt{x} being \texttt{4}, the following trace is
produced:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(x~==~4)\\
y~=~y~+~x
}
\end{quote}

@@

There are cases in which it is useful to turn an arbitrary variable
into a constant value. This process is called \emph{promotion} and it is an old idea
in partial evaluation (it's called ``the trick'' there). Promotion is also heavily
used by \href{http://psyco.sourceforge.net/}{Psyco} and by all older versions of PyPy's JIT. Promotion is a technique
that only works well in JIT compilers, in
static compilers it is significantly less applicable.

@@

We get a trace that looks like this:
%
\begin{quote}{\ttfamily \raggedright \noindent
v1~=~x~*~2\\
z~=~v1~+~1\\
v2~=~z~+~y\\
return(v2)
}
\end{quote}

@@

The meaning of this hint is that the tracer should pretend that \texttt{x} is a
constant
in the code that follows. When just running the code, the function has no
effect, as it simply returns its first argument. When tracing, some extra work
is done. Let's assume that this changed function is traced with
the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
operation at the beginning:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(x~==~4)\\
v1~=~x~*~2\\
z~=~v1~+~1\\
v2~=~z~+~y\\
return(v2)
}
\end{quote}

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the value of \texttt{x} as it was at runtime. From the point of view of the
optimizer, this guard is not any different than the one produced by the \texttt{if}
statement in the example above. After the guard, the rest of the trace can
assume that \texttt{x} is equal to \texttt{4}, meaning that the optimizer will turn this
trace into:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(x~==~4)\\
v2~=~9~+~y\\
return(v2)
}
\end{quote}

@@

If this trace is executed with values of \texttt{x} other than \texttt{4}, the guard will
fail, and execution will continue in the interpreter. If the guard fails often
enough, a new trace will be started from the guard. This other trace will
capture a different value of \texttt{x}. If it is e.g. \texttt{2}, then the optimized
trace looks like this:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(x~==~2)\\
v2~=~5~+~y\\
return(v2)
}
\end{quote}

@@

Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
trace (note how the call to \texttt{compute} is inlined):
%
\begin{quote}{\ttfamily \raggedright \noindent
x~=~a.x\\
v1~=~x~*~2\\
v2~=~v1~+~1\\
v3~=~v2~+~val\\
a.y~=~v3
}
\end{quote}

@@

Now the trace will look like this:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(a~==~0xb73984a8)\\
v1~=~compute(a)\\
v2~=~v1~+~val\\
a.y~=~v2
}
\end{quote}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since \texttt{compute} function is marked as pure, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(a~==~0xb73984a8)\\
v2~=~9~+~val\\
a.y~=~v2
}
\end{quote}

@@

With this changed instance implementation, the trace we had above changes to the
following, where \texttt{0xb74af4a8} is the memory address of the Map instance that
has been promoted, see Figure~\ref{fig:trace2}.

@@

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
a pure function and they have constant arguments. That means that \texttt{index1/2/3}
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. The optimized trace looks can be seen in Figure~\ref{fig:trace3}

\begin{figure}
\input{code/trace3.tex}
\caption{Optimized Trace After the Introduction of Maps}
\label{fig:trace3}
\end{figure}

The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
of the constant-folded \texttt{getindex} call. This trace is already much better than
the original one. Now we are down from five dictionary lookups to just two.

@@

The constants \texttt{41} and \texttt{17} are the results of the folding of the
\texttt{\_find\_method`} calls. This final trace is now very good. It no longer performs any
dictionary lookups. Instead it contains several guards. The first guard
checks that the map is still the same. This guard will fail if the same
code is executed with an instance that has another layout. The second guard
checks that the class of \texttt{inst} is still the same. It will fail if trace is
executed with an instance of another class. The third guard checks that the
class did not change since the trace was produced. It will fail if somebody
calls the \texttt{change\_method} method on the class.

>>>>

\definecolor{gray}{rgb}{0.5,0.5,0.5}

@@

\section{Background}
\label{sec:Background}

\subsection{The PyPy Project}
\label{sect:pypy}

PyPy's approach to implementing dynamic
languages is to write an interpreter for
the language in RPython. This interpreter can be translated to C and then
further to machine code. The interpreter consists of code in the form of a
large number of generated C functions and some data. Similarly, the user
program consists of functions in the language the interpreter executes.

@@

%___________________________________________________________________________
\subsection{PyPy's Meta-Tracing JIT Compilers}
\label{sect:tracing}

PyPy's JIT is a meta-tracer \cite{bolz_tracing_2009}. Since we want to re-use
our tracer for a variety of languages, we
don't trace the execution of the user program, but instead trace the execution
of the \emph{interpreter} that is running the program. This means that the traces
don't contain the bytecodes of the language in question, but RPython-level
operations that the interpreter did to execute the program.

@@

\subsection{Optimizing Traces}
\label{sub:optimizing}

Before sending the trace to the backend to produce actual machine code, it is
optimized.  The optimizer applies a number of techniques to remove or reduce
the number of operations: most of these are well known compiler optimization
techniques, with the difference that it is easier to apply them in a tracing
JIT because it only has to deal with linear traces.  Among the techniques:
%
\begin{itemize}
    \item constant folding
    \item common subexpression elimination
    \item allocation removal \cite{bolz_allocation_2011}
    \item store/load propagation
    \item loop invariant code motion
\end{itemize}

% section Background (end)
%___________________________________________________________________________

@@

In this section we will describe how to add two hints that allow the
interpreter author to increase the optimization opportunities for constant
folding. If applied correctly these techniques can give really big speedups by
pre-computing parts of what happens at runtime. On the other
hand, if applied incorrectly they might lead to code bloat, thus making the
resulting program actually slower.

For constant folding to work, two conditions need to be met:

\begin{itemize}
    \item the arguments of an operation actually need to all be constant,
    i.e. statically known by the optimizer
    \item the operation needs to be \emph{pure}, i.e. always yield the same result given
    the same arguments.
\end{itemize}

@@

@@

If the fragment is traced with \texttt{x} being \texttt{4}, the following trace is
produced:
%
\begin{Verbatim}
guard(x == 4)
y = y + x
\end{Verbatim}

@@

There are cases in which it is useful to turn an arbitrary variable
into a constant value. This process is called \emph{promotion} and it is an old idea
in partial evaluation (it's called ``the trick'' \cite{XXX} there). Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} and by all older versions
of PyPy's JIT. Promotion is a technique that only works well in JIT compilers,
in static compilers it is significantly less applicable.

@@

We get a trace that looks like this:

\begin{Verbatim}
v1 = x * 2
z = v1 + 1
v2 = z + y
return(v2)
\end{Verbatim}

@@

The meaning of this hint is that the tracer should pretend that \texttt{x} is a
constant
in the code that follows. When just running the code, the function has no
effect, as it simply returns its first argument. When tracing, some extra work
is done. Let's assume that this changed function is traced with
the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
operation at the beginning:

\begin{Verbatim}
guard(x == 4)
v1 = x * 2
z = v1 + 1
v2 = z + y
return(v2)
\end{Verbatim}

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the value of \texttt{x} as it was at runtime. From the point of view of the
optimizer, this guard is not any different than the one produced by the \texttt{if}
statement in the example above. After the guard, the rest of the trace can
assume that \texttt{x} is equal to \texttt{4}, meaning that the optimizer will turn this
trace into:

\begin{Verbatim}
guard(x == 4)
v2 = 9 + y
return(v2)
\end{Verbatim}

@@

If this trace is executed with values of \texttt{x} other than \texttt{4}, the guard will
fail, and execution will continue in the interpreter. If the guard fails often
enough, a new trace will be started from the guard. This other trace will
capture a different value of \texttt{x}. If it is e.g. \texttt{2}, then the optimized
trace looks like this:

\begin{Verbatim}
guard(x == 2)
v2 = 5 + y
return(v2)
\end{Verbatim}

@@

Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
trace (note how the call to \texttt{compute} is inlined):
%
\begin{Verbatim}
x = a.x
v1 = x * 2
v2 = v1 + 1
v3 = v2 + val
a.y = v3
\end{Verbatim}

@@

Now the trace will look like this:
%
\begin{Verbatim}
guard(a == 0xb73984a8)
v1 = compute(a)
v2 = v1 + val
a.y = v2
\end{Verbatim}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since \texttt{compute} function is marked as pure, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{Verbatim}
guard(a == 0xb73984a8)
v2 = 9 + val
a.y = v2
\end{Verbatim}

@@

With this changed instance implementation, the trace we had above changes to the
following, where \texttt{0xb74af4a8} is the memory address of the Map instance that
has been promoted, see Figure~\ref{fig:trace2}. Operations that can be
optimized away are grayed out.

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
a pure function and they have constant arguments. That means that \texttt{index1/2/3}
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. This trace is already much better than
the original one. Now we are down from five dictionary lookups to just two.

@@

@@

The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
of the constant-folded \texttt{getindex} call.
The constants \texttt{41} and \texttt{17} are the results of the folding of the
\texttt{\_find\_method`} calls. This final trace is now very good. It no longer performs any
dictionary lookups. Instead it contains several guards. The first guard
checks that the map is still the same. This guard will fail if the same
code is executed with an instance that has another layout. The second guard
checks that the class of \texttt{inst} is still the same. It will fail if trace is
executed with an instance of another class. The third guard checks that the
class did not change since the trace was produced. It will fail if somebody
calls the \texttt{change\_method} method on the class.
